---
title: "Differential Gene Expression using computational dissection of our single cell data"
author: "Alberto Perez-Posada"
date: "2024-12-06"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
dir <- "C:/Users/sophie.peron/OneDrive - University of Exeter/Cell Cycle G1G2 CDH1 paper shared folder/Figure Jupyter Notebooks/h5ad/analysis_250528_f/DGE/Cdh1"
#dir <- "C:/Users/sp1046/OneDrive - University of Exeter/Cell Cycle G1G2 CDH1 paper shared folder/Figure Jupyter Notebooks/h5ad/analysis_250528_f/DGE/Cdh1"
setwd(dir)
fcha <- function(){ gsub("-","",Sys.Date()) }

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = dir)
options(scipen=999)
Sys.setenv(VROOM_CONNECTION_SIZE=5000072)
```

## About

Differential gene expression analysis after RNAi treatment: aggregation of tte data into pseudobulk counts, differential gene expression analysis at the level of the broad groups using DESeq

## Load libraries

```{r libraries, message=FALSE, warning=FALSE}
require(ComplexHeatmap)
library(colorspace)
library(cowplot)
require(DESeq2)
require(dplyr)
library(eulerr)
require(ggplot2)
library(ggrepel)
library(ggrastr)
library(gridExtra)
library(readr)
library(Matrix)
```

## Load functions

We use here a wrapper of DESeq and a couple of plotting functions:

 - It takes the a count matrix and a samples information matrix to perform DGE analysis.
 - If no DGEs are found, it ends with a message indicating so.
 - If it only finds one DGE, it reports its ID.
 - If it finds more than one DGE, it proceeds to plot a volcano plot and a heatmap of normalised gene expression.
 - In the two latter cases, a list of things is returned: the DESeq2 object which contains the whole DGE analysis, the results table from running DESeq2, and a list of differentially expressed genes.

```{r load_functions}
source("../../Common R Functions/sourcefolder.R")
sourceFolder("../../Common R Functions/", recursive = TRUE)
```

## Load data

```{r load_data}

# Load metadata
Idents <- read.delim2("rnai_cells_metadata_for_DGE.csv",
sep = "," , header = TRUE)

# Load the matrix
matrix_X <- t(readMM(file = "rnai_raw_for_DGE.mtx.gz"))
colnames(matrix_X) <- read.csv("rnai_cells_metadata_for_DGE.csv")[,1]
rownames(matrix_X) <- read.csv("rnai_all_genes_metadata_for_DGE.csv")[,1]
matrix_X <- matrix_X[,colnames(matrix_X) %in% Idents$X]

```

## Pseudobulk at the cluster/condition/replicate intersection

### Count matrix

We retrieve a pseudobulk matrix by taking into account the cell type, the replicate and experimental condition. This will effectively retrieve a matrix of "pseudo-samples", where a sample is any combination of the following: "cell type" + "experimental condition" + "replicate" .

We separate the cells in 3 categories based on the neoblast score.

```{r}
Idents <- Idents %>%
  mutate(
    cell_category = case_when(
      neoblast_score > 0.259 ~ "neoblasts",
      neoblast_score <= 0.259 & neoblast_score > 0.171 ~ "progenitors",
      neoblast_score < 0.171 ~ "differentiated"
    ))

```


```{r psbulk_broad}
# run pseudobulk
scdata_metadata <- Idents
cdh_psbulk_cond_rep_broad <- pseudobulk_cond_rep(
  x = matrix_X,
  identities = scdata_metadata$cell_category,
  conditions = scdata_metadata$Condition,
  replicates = scdata_metadata$Replicate
)
```


```{r}
# check the count matrix
cdh_psbulk_cond_rep_broad$matrix[1:5,1:5]
```

```{r}
dim(cdh_psbulk_cond_rep_broad$matrix)
```

### Sample Information table

This table help providing DESeq2 with all the relevant information for the samples.
The layout is something like this:

```
sample condition replicate
S01    Cdh1     1
S12    GFP      1
S23    Cdh1     2
S34    GFP      2
```

Every row corresponds to a column in the counts matrix and it has the descriptions of what they are.

Column "sample" must be the names of the columns in the count matrix.
necessary columns: 'condition' with the identity of the experiment, and  'replicate' with the identity of the replicate of the experiment.

```{r parse_sampletable}
clean_sampletable <- function(s){
  s <- s[s$condition != "",]
  s <-
    s[
      s$ctype %in%
        names(table(s$ctype))[ # those cell types
          table(s$ctype) == 
            (length(unique(s$condition))*length(unique(s$replicate))) # possible combinations of condition x replicates 
        ],
    ]
  s$condition <- factor(s$condition, levels = unique(s$condition))
  
  return(s)
}

# Broad cell type
cdh_psbulk_cond_rep_broad$sampletable <- clean_sampletable(cdh_psbulk_cond_rep_broad$sampletable)
head(cdh_psbulk_cond_rep_broad$sampletable)
```



```{r}
cdh_sampletable_broad <- cdh_psbulk_cond_rep_broad$sampletable
cdh_sampletable_broad$ctype <- 
  factor(
    cdh_sampletable_broad$ctype,
    levels = unique(cdh_sampletable_broad$ctype)
    )

cdh_sampletable_broad$condition <- factor(cdh_sampletable_broad$condition, levels = c("Cdh1","H2B", "GFP"))
cdh_sampletable_broad$replicate <- factor(cdh_sampletable_broad$replicate, levels = c("1","2"))
head(cdh_sampletable_broad)
```

Filtering to exclude samples missing either a cell type, a condition or a replicate

```{r}
cdh_matrix_broad <- 
  cdh_psbulk_cond_rep_broad$matrix[
    ,
    colnames(cdh_psbulk_cond_rep_broad$matrix) %in% cdh_sampletable_broad$sample
  ]
```

## Analysis

Here we run a massive DESeqwrapper for all cell types.

This is a small wrapper of a much larger function called `deseq_sc`. 

The small wrapper iterates over cell types, parses the sampletable and the expression matrix, and runs deseq for every cell type separately.

The larger wrapper, `deseq_sc`, runs the following:

 - It simplifies the sampletable data and grabs just the key minimal information to create a design matrix (i.e. condition, replicate, ...)
 - It parses the information of the contrast and creates a formula accordingly
 - It creates a DESeq2 object with this information plus the expression matrix supplied to the function (which contains the relevant columns in relation to the sample, condition, replicates ... indicated in the sampletable).
 - It filters the genes with less than a minimum number of counts in a minimum number of the samples.
 - If specified (keep_dubious), it keeps the genes that have a minimum number of counts even if they don't pass the criteria above.
 - It runs the DESeq2 wrapper with all the relevant parameters provided by the user.
 - It retrieves the results using the contrast provided by the user.
 - It filters the results based on p-value or p-adj, following the user's indication, as well as the max. value (0.05, 0.1, ...)
 - It builds a list of output results.


```{r}
cdh_DGE_all_broad <- list()
for(i in unique(cdh_psbulk_cond_rep_broad$sampletable$ctype)){
  message("starting with cell type ",i)
  cdh_DGE_all_broad[[i]] <-
    deseq_pseudobulk(
      count_matrix = cdh_matrix_broad,
      samples_info = cdh_sampletable_broad[,-2],
      celltype = i,
      #filter_by = "pvalue", 
      p_threshold = 0.05,
      contrast_info = c("condition","Cdh1","GFP"), # change for control here
      plot_results = FALSE, min_passing_samples = 3,
      min_counts_per_sample = 3,
      keep_dubious = FALSE
    )
  message("done cell type ",i)
}
```

## Metaplot (broad)

With this plot we will show which cell types have the highest amount of D.E.G.s.

As a control we can plot this compared to the cluster size, because one expects that cell types with better coverage will capture more nuances in gene expression. 

```{r}
cdh_diffreg_broad <- data.frame(
  num_diff = sapply(
    cdh_DGE_all_broad,
    function(x){
      a = x$diffgenes
      if(is.na(a[1])){
        b = 0
      } else{
        b = length(a)
      }
      return(b)
    }
  ),
  total_genes = sapply(
    cdh_DGE_all_broad, function(x){nrow(x$res)}
    )
)
```


```{r}
# fraction of differentially expressed genes
cdh_diffreg_broad$fraction_diff <- 
  cdh_diffreg_broad$num_diff / cdh_diffreg_broad$total_genes
```


```{r}
# number of cells per cluster 
cdh_diffreg_broad <-
  merge(
    cdh_diffreg_broad,
    table(scdata_metadata$cell_category),
    by.x = 0,
    by.y = 1,
    all.x = TRUE
  )
colnames(cdh_diffreg_broad)[1] <- "cell_category"
```


We will quickly calculate the pseudobulk at the broadtype level (without replicates or anything) to retrieve the counts for Cdh1 for later plotting.

```{r}
smed_cdh_counts_broad <- pseudobulk(x=matrix_X,identities = scdata_metadata$cell_category)
```

```{r}
# cdh1 h1SMcG0005288 FZR1 in rosetta
cdh <- "h1SMcG0005288" #cdh1
#cdh <- "h1SMcG0016295" #h2b
cdh_diffreg_broad <-
  merge(
    cdh_diffreg_broad,
    data.frame(ctype = colnames(smed_cdh_counts_broad), cdh = quantnorm(smed_cdh_counts_broad)[rownames(smed_cdh_counts_broad)==cdh,]),
    by = 1
  )
```

Here is the plot of the number of diff.reg. genes (logarithmic) and the level of expression of Cdh1 (logarithmic, quantile-normalised)

```{r, fig.width = 5, fig.height = 5}
p1 <- ggplot(cdh_diffreg_broad, aes(x = log1p(cdh), y = log1p(num_diff), label = cell_category)) +
  geom_point() +
  labs(
    title = "Cdh1 expression vs # DiffReg",
    x = "Cdh1 expression (log, quan.norm)",
    y = "log no. diffreg genes (pval<.05)"
  ) +
  scale_color_identity()+
  theme_minimal()+
  geom_text_repel(cex=3) + labs(title = "Cdh1 expression vs # DiffReg")
p1
```


We can also check this in relation to the size of the cluster. 

```{r, fig.width = 5, fig.height = 5}
p2 <- ggplot(cdh_diffreg_broad, aes(x = log(Freq), y = log1p(num_diff), label = cell_category)) +
  geom_point() +
  labs(
    title = "broad cluster size vs # DiffReg",
    x = "broad cluster size (log(freq))",
    y = "num. diffreg genes (pval<.05)"
  ) +
  scale_color_identity()+
  theme_minimal()+
  geom_text_repel(cex=3) + labs(title = "broad cluster size vs # DiffReg")

p2
```

## Volcano plots

```{r, fig.width = 5, fig.height = 3}
y_max <- 60 # to homogenise axis

list_volcanos <- list() # list to store the plots

# create a list with names(cdh_DGE_all_broad) to choose the order of the plots
names_cdh_DGE_all_broad <- list("neoblasts",  "progenitors" ,   "differentiated")

for(celltype in names_cdh_DGE_all_broad){
  list_volcanos[[celltype]] <- cdh_DGE_all_broad[[celltype]]$res %>%
    ggplot(aes(x = log2FoldChange, y = -log(pvalue), color = factor(padj < 0.05))) +
    geom_point_rast(size = .75) +
    scale_color_manual(values = c("FALSE" = unname(gb_col["black1"]), "TRUE" = "#e66056")) +
    ylim(c(0,y_max))+
    guides(color = "none") + 
    theme_classic() +
    theme(panel.border = element_rect(colour = "black", fill=NA, linewidth = 0.75), axis.title.x=element_blank(), axis.title.y=element_blank())+
    ggtitle(label = celltype)  + 
   # ggtitle(label = paste(strwrap(celltype, width = 15), collapse = "\n"))  + 
    theme(text=element_text(size=8))
}

do.call("grid.arrange", c(list_volcanos, ncol = 3))

```

## save the data

```{r}
#saves the list of differentially expressed genes and the lfc in csv files for further analysis using python
# identitity of the DE genes are stored here : cdh_DGE_all_broad[["neoblasts"]][["diffgenes"]]
# lfcs are stored here: cdh_DGE_all_broad[["neoblasts"]][["res"]]
getwd()

for (i in names(cdh_DGE_all_broad)){
  print(i)
  li <- cdh_DGE_all_broad[[i]][["diffgenes"]]
  tablelfc <- cdh_DGE_all_broad[[i]][["res"]]
  tablelfc$deg = ifelse(tablelfc$padj < 0.05, "DEG", "none") # add a column with DGE results
  
  i <- gsub("/", "", i)
  write.table(tablelfc, file = paste0('./outputs_V2scores/', i, '_DGE_res_Cdh1.tsv'),  sep = "\t", quote= FALSE, col.names = TRUE )
}

```

```{r}
save(
  cdh_DGE_all_broad,
  cdh_diffreg_broad,
  cdh_matrix_broad,
  cdh_sampletable_broad,
  Idents,
  matrix_X,
  smed_cdh_counts_broad,
  file = "./outputs_V2scores/260304_Cdh1_diffexp.rda"
)
```

## repeat for H2B

! this will overwrite the cdh1 variables

```{r}
cdh_DGE_all_broad <- list()
for(i in unique(cdh_psbulk_cond_rep_broad$sampletable$ctype)){
  message("starting with cell type ",i)
  cdh_DGE_all_broad[[i]] <-
    deseq_pseudobulk(
      count_matrix = cdh_matrix_broad,
      samples_info = cdh_sampletable_broad[,-2],
      celltype = i,
      #filter_by = "pvalue", 
      p_threshold = 0.05,
      contrast_info = c("condition","H2B","GFP"), # change for control here
      plot_results = FALSE, min_passing_samples = 3,
      min_counts_per_sample = 3,
      keep_dubious = FALSE
    )
  message("done cell type ",i)
}
```

## Metaplot (broad)

With this plot we will show which cell types have the highest amount of D.E.G.s.

As a control we can plot this compared to the cluster size, because one expects that cell types with better coverage will capture more nuances in gene expression. 

```{r}
cdh_diffreg_broad <- data.frame(
  num_diff = sapply(
    cdh_DGE_all_broad,
    function(x){
      a = x$diffgenes
      if(is.na(a[1])){
        b = 0
      } else{
        b = length(a)
      }
      return(b)
    }
  ),
  total_genes = sapply(
    cdh_DGE_all_broad, function(x){nrow(x$res)}
    )
)
```


```{r}
# fraction of differentially expressed genes
cdh_diffreg_broad$fraction_diff <- 
  cdh_diffreg_broad$num_diff / cdh_diffreg_broad$total_genes
```


```{r}
# number of cells per cluster 
cdh_diffreg_broad <-
  merge(
    cdh_diffreg_broad,
    table(scdata_metadata$cell_category),
    by.x = 0,
    by.y = 1,
    all.x = TRUE
  )
colnames(cdh_diffreg_broad)[1] <- "cell_category"
```

We will quickly calculate the pseudobulk at the broadtype level (without replicates or anything) to retrieve the counts for Cdh1 for later plotting.

```{r}
smed_cdh_counts_broad <- pseudobulk(x=matrix_X,identities = scdata_metadata$cell_category)
```

```{r}
# cdh1 h1SMcG0005288 FZR1 in rosetta
#cdh <- "h1SMcG0005288" #cdh1
cdh <- "h1SMcG0016295" #h2b
cdh_diffreg_broad <-
  merge(
    cdh_diffreg_broad,
    data.frame(ctype = colnames(smed_cdh_counts_broad), cdh = quantnorm(smed_cdh_counts_broad)[rownames(smed_cdh_counts_broad)==cdh,]),
    by = 1
  )
```

Here is the plot of the number of diff.reg. genes (logarithmic) and the level of expression of Cdh1 (logarithmic, quantile-normalised)

```{r, fig.width = 5, fig.height = 5}
p1 <- ggplot(cdh_diffreg_broad, aes(x = log1p(cdh), y = log1p(num_diff), label = cell_category)) +
  geom_point() +
  labs(
    title = "H2B expression vs # DiffReg",
    x = "H2B expression (log, quan.norm)",
    y = "log no. diffreg genes (pval<.05)"
  ) +
  scale_color_identity()+
  theme_minimal()+
  geom_text_repel(cex=3) + labs(title = "H2B expression vs # DiffReg")
p1
```


We can also check this in relation to the size of the cluster. 

```{r, fig.width = 5, fig.height = 5}
p2 <- ggplot(cdh_diffreg_broad, aes(x = log(Freq), y = log1p(num_diff), label = cell_category)) +
  geom_point() +
  labs(
    title = "broad cluster size vs # DiffReg",
    x = "broad cluster size (log(freq))",
    y = "num. diffreg genes (pval<.05)"
  ) +
  scale_color_identity()+
  theme_minimal()+
  geom_text_repel(cex=3) + labs(title = "broad cluster size vs # DiffReg")

p2
```

## Volcano plots

```{r, fig.width = 5, fig.height = 3}
y_max <- 60 # to homogenise axis

list_volcanos <- list() # list to store the plots

# create a list with names(cdh_DGE_all_broad) to choose the order of the plots
names_cdh_DGE_all_broad <- c("neoblasts",  "progenitors" ,   "differentiated")

for(celltype in names_cdh_DGE_all_broad){
  list_volcanos[[celltype]] <- cdh_DGE_all_broad[[celltype]]$res %>%
    ggplot(aes(x = log2FoldChange, y = -log(pvalue), color = factor(padj < 0.05))) +
    geom_point_rast(size = .75) +
    scale_color_manual(values = c("FALSE" = unname(gb_col["black1"]), "TRUE" = "#e66056")) +
    ylim(c(0,y_max))+
    guides(color = "none") + 
    theme_classic() +
    theme(panel.border = element_rect(colour = "black", fill=NA, linewidth = 0.75), axis.title.x=element_blank(), axis.title.y=element_blank())+
    ggtitle(label = celltype)  + 
   # ggtitle(label = paste(strwrap(celltype, width = 15), collapse = "\n"))  + 
    theme(text=element_text(size=8))
}


do.call("grid.arrange", c(list_volcanos, ncol = 3))

```

## save the data

```{r}
#saves the list of differentially expressed genes and the lfc in csv files for further analysis using python
# identitity of the DE genes are stored here : cdh_DGE_all_broad[["neoblasts"]][["diffgenes"]]
# lfcs are stored here: cdh_DGE_all_broad[["neoblasts"]][["res"]]
getwd()

for (i in names(cdh_DGE_all_broad)){
  print(i)
  li <- cdh_DGE_all_broad[[i]][["diffgenes"]]
  tablelfc <- cdh_DGE_all_broad[[i]][["res"]]
  tablelfc$deg = ifelse(tablelfc$padj < 0.05, "DEG", "none") # add a column with DGE results
  
  i <- gsub("/", "", i)
  write.table(tablelfc, file = paste0('./outputs_V2scores/', i, '_DGE_res_H2B.tsv'),  sep = "\t", quote= FALSE, col.names = TRUE )
}

```

```{r}
save(
  cdh_DGE_all_broad,
  cdh_diffreg_broad,
  cdh_matrix_broad,
  cdh_sampletable_broad,
  Idents,
  matrix_X,
  smed_cdh_counts_broad,
  file = "./outputs_V2scores/260304_H2B_diffexp.rda"
)
```



## The End!

```{r}
sessionInfo()
```